# Week 4, Class 2: Network Analysis with R"


## Introduction  


## Constructing a Network in R  

One of the easiest data formats to construct a network is an edge list: a simple dataframe with two columns, representing the connections between two nodes, one per row. It makes particular sense with correspondence data, which is often stored as a table of letters with a 'from' and a 'to'---more or less a ready-made edge list. In a correspondence dataset you might also have multiple sets of each of the edges (multiple letters between the same pair of individuals). You can add this to the edge list as an attribute called weight, which is simply another column.

I use three R network libraries to do almost everything network-related, from analysis to visualisation: ```igraph```, ```tidygraph``` and ```ggraph```. My goal is to port everything to a format which is really easy to work with using existing my data analysis workflow. That format is known as 'tidy data', and it is a way of working with data which is easily transferable across a range of uses. It also means you need to learn very little new programming to do network analysis if you stay within this 'ecosystem'. 

### Import Network Data with the Tidyverse

The whole workflow uses four packages. The first is ```tidyverse```, a collection of various packages used for data wrangling and analytics.

```{r message=FALSE, warning=FALSE}

library(dplyr)
library(magrittr)
library(readr)
library(stringr)

```

Next I need some network data. On the project this is generally a comma separated values file containing the information for one letter per row. We turn the raw letter data into an edge list in a standardised format, with the from/to information, a letter date, a place, and a unique identifier for the letter. Because the dataset is large and the same names are often used repeatedly, we use a unique ID number rather than people or place names when constructing the edge list. The nice thing about working within the tidyverse is that it's easy to then match the IDs back to actual names afterwards. 

These edge lists can be turned into a network object using ```tidygraph``` and the function ```as_tbl_graph```. This function takes the first two columns in a dataset and turns them into the network graph, using any additional columns as attributes. We use two separate tables of data. First, an edge list, using unique, unambiguous numeric IDs for people and place names. Next, additional lookup tables, just with the unique ID, the person's name, and additional information, if you have it (which can be used to filter the network afterwards). 

First load the two tables into R (I've created small sample tables to work with), which are available [here](https://raw.githubusercontent.com/networkingarchives/blog/master/2021/04/15/my-network-analysis-workflow/letters.csv) and [here](https://raw.githubusercontent.com/networkingarchives/blog/master/2021/04/15/my-network-analysis-workflow/letters.csv): 

```{r}
letters = read_csv("https://raw.githubusercontent.com/yann-ryan/dh_intro_gates/main/sample_letter_data_network.csv", col_types = cols(.default = "c"))

```


If you have multiple letters between individuals, you can sum them and use as a weight in the network, or you can ignore it. You can do this with ```tidyverse``` commands: ```group_by()``` and ```tally()```, changing the name of the new column to 'weight'. 

```{r message=FALSE, warning=FALSE}

edge_list = letters %>% 
  group_by(author, recipient) %>% 
  tally(name = 'weight')

```

### Turn the edge list into a tbl_graph

Next transform the edge list into a network object called a tbl_graph, using [tidygraph](https://www.data-imaginist.com/2017/introducing-tidygraph/). A tbl_graph is a graph object which can be manipulated using tidyverse grammar. This means you can create a network and then use a range of standard data analysis functions on it as needed, without learning a whole new set of commands.

First, load the tidygraph library. Use ```as_tbl_graph()``` to turn the edge list into a network. The first two columns will be taken as the from and to data, and any additional columns added as attributes. It'll automatically create a nodes table, too.

```{r message=FALSE, warning=FALSE}
library(tidygraph)

sample_tbl_graph = edge_list %>% 
  as_tbl_graph()

sample_tbl_graph
```


The tbl_graph is an object with two tables, one for the edges and one for the nodes. You can access each of the tables using the function activate(nodes) or activate(edges). The active table has (active) after it. 

```{r}

sample_tbl_graph = sample_tbl_graph %>% 
  activate(edges)

sample_tbl_graph

```

### Calculate network metrics 

Global metrics

Don't need tidygraph (but doesn't matter if it is)

#### Node-level metrics with a table.

Tidygraph allows you to perform calculations on the tbl_graph using ```mutate```, using standard igraph algorithms. So for example to calculate the degree of every node:

```{r}
sample_tbl_graph %>% 
  activate(nodes) %>% 
  mutate(degree = centrality_degree(mode = 'total'))
```
If you run standard functions meant to be used on a dataframe, they will happen to the active table. So if you wanted to filter just edges from ID 1, for example, you could use the ```filter``` verb from dplyr:


```{r}
sample_tbl_graph %>% filter(from ==1)
```


Maybe you only want to keep nodes with the title 'Sir'? Activate the nodes again with ```activate(nodes)```, join the people table, then use ```filter``` and ```str_detect``` to filter based on a regular expressions pattern. You'll see that it has filtered out unused edges now, too:


```{r}
sample_tbl_graph %>% 
  activate(nodes)%>% 
  filter(str_detect(name, "(?i)sir"))
```

Slightly more useful might be to filter based on some calculation you've made previously. The data format allows you to use dplyr pipes (%>%) to perform one calculation on the data, then pass that new dataframe along to the next function. It works really well with tidygraph. Here we calculate the degree scores first, then filter to include only nodes with a degree score over two:

```{r}

sample_tbl_graph %>% 
  activate(nodes) %>% 
  mutate(degree = centrality_degree(mode = 'total')) %>% 
  filter(degree >2)

```


## Exercises:

Load in the network you created the previous week.

Create a code chunk below to find the node with the highest degree in your network









